\thetitle

### 2020-11-08 sun_phases/spin_qudit_tomo.tex
further tightened the error bound

Consider a $d$-state spin qudit with total spin $s\equiv\frac{d-1}{2}$.
The defining property of a spin qudit is the fact that it describes an angular momentum degree of freedom, which has specific implications for how a spin qudit should transform under the group SO(3) of rotations in 3D space.
Due to the central importance of these transformation rules for a spin qudit, we first identify a basis of operators that transforms nicely under 3D rotations.
One such basis is provided by the {\it transition operators} (also known as polarization operators \cite{kryszewski2006positivity, bertlmann2008bloch}), defined by
\begin{align}
  T_{\ell m} \equiv \sqrt{\f{2\ell+1}{2s+1}} \sum_{\mu,\nu=-s}^s
  \bk{s\mu;\ell m|s\nu} \op{\nu}{\mu},
  \label{eq:trans_op}
\end{align}
where $\bk{s\mu;\ell m|s\nu}$ is a Clebsh-Gordan coefficient that enforces $\ell\in\set{0,1,\cdots,2s}$ and $m\in\set{-\ell,-\ell+1,\cdots,\ell}$, such that there are $d^2$ operators in total.
For brevity, we will generally treat the value of $d$ (or $s$) as constant but arbitrary throughout this work, and suppress any explicit dependence of quantities or operators such as $T_{\ell m}$ on $d$.
The transition operators are orthonormal with respect to the trace inner product, and transform nicely under conjugation:
\begin{align}
  \obk{T_{\ell m }|T_{\ell'm'}}
  = \delta_{\ell\ell'} \delta_{mm'},
  &&
  T_{\ell m}^\dag = \p{-1}^m T_{\ell,-m},
\end{align}
where for any $d\times d$ matrix $X=\sum_{\mu,\nu} X_{\mu\nu} \op{\mu}{\nu}$ we define the $d^2$-component vector $\oket{X} \equiv \sum_{\mu,\nu} X_{\mu\nu} \ket{\mu\nu}$; $\obra{X}$ is the conjugate transpose of $\oket{X}$; and $\delta_{kk'}\equiv 1$ if $k=k'$ and $0$ otherwise.
As a consequence, we can expand any density operator $\rho$ in the transition operator basis as
\begin{align}
  \rho = \sum_{\ell=0}^{2s} \sum_{m=-\ell}^\ell \rho_{\ell m} T_{\ell m},
  &&
  \rho_{\ell m} \equiv \obk{T_{\ell m}|\rho},
  \label{eq:trans_state}
\end{align}
where $\rho^\dag=\rho$ implies that $\rho_{\ell m}^*=\p{-1}^m\rho_{\ell,-m}$.
The transition operators can be physically interpreted as the transition $\ket\psi\to T_{\ell m}\ket\psi$ induced on the state $\ket\psi$ of a spin-$s$ system by the absorption of a spin-$\ell$ particle with spin projection $m$ onto a fixed quantization axis.
In analogy with the spherical harmonics $Y_{\ell m}$, we will refer to the index $\ell$ as the {\it degree} and $m$ as the {\it order} of $T_{\ell m}$.

@@

Rather than the statistical errors $\epsilon_{i\mu} \equiv \tilde\Pi_{i\mu} - \bk{\Pi_{i\mu}}_\rho$ in the estimates $\tilde\Pi_{i\mu}$ of the projectors $\Pi_{i\mu}$, we now consider the statistical errors $\epsilon_{i\ell} \equiv \tilde T_{i\ell} - \bk{T_{i\ell}}_\rho$ in the estimates $\tilde T_{i\ell}$ of the transition operators $T_{i\ell} \equiv T_{\v v_i\ell,0}$.
We can expand the transition operators $T_{i\ell}$ as a sum over projectors $\Pi_{i\mu}$ as
\begin{align}
  T_{i\ell} = \sum_\mu t_{\ell\mu} \Pi_{i\mu},
  &&
  t_{\ell\mu} \equiv \bk{\mu|T_{\ell,0}|\mu},
\end{align}
and likewise $\tilde T_{i\ell} = \sum_\mu t_{\ell\mu} \tilde\Pi_{i\mu}$.
The covariances between statistical errors in the transition operator basis are then
\begin{align}
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell'}}
  = \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \bbk{\epsilon_{i\mu} \epsilon_{i'\mu'}}
  = \delta_{ii'} \times \f1n \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \sp{\delta_{\mu\mu'} p_\mu^{(i)} - p_\mu^{(i)} p_{\mu'}^{(i)}}.
\end{align}
In order to establish an upper bound on reconstruction error for an unknown quantum state, we need an upper bound on the variance $\bbk{\epsilon_{i\ell}^2}$ in particular.
To this end, for any multinomial probability distribution $p\equiv\p{p_1,p_2,\cdots,p_d}$ we define the function
\begin{align}
  g_\ell\p{p} \equiv \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell\mu'}
  \p{\delta_{\mu\mu'} p_\mu - p_\mu p_{\mu'}}
  = \sum_\mu t_{\ell\mu}^2 p_\mu - \p{\sum_\mu t_{\ell\mu} p_\mu}^2.
\end{align}
Ideally, we would maximize this function over all valid probability distributions, but doing so is rather difficult.
Instead, we can find a loose upper bound by ignoring the negative terms:
\begin{align}
  g_\ell\p{p} \le \sum_\mu t_{\ell\mu}^2 p_\mu
  \le \max_\mu t_{\ell\mu}^2
  = \f{2\ell+1}{d} \max_\mu \bk{s\mu;\ell,0|s\mu}^2
  \le \f{2\ell+1}{d},
\end{align}
which implies that
\begin{align}
  \bbk{\epsilon_{i\ell}^2} = \f1n g_\ell\p{p^{(i)}} \le \f{2\ell+1}{nd}.
  \label{eq:variance_bound}
\end{align}
Note that $t_{0,\mu}=1/\sqrt{d}$, so $f_0\p{p}=0$ and $\bbk{\epsilon_{i,0}^2}=0$.

@@

Equipped with the bound on variances in Eq.~\eqref{eq:variance_bound}, we again consider the block-diagonalized form of the measurement matrix derived in Section \ref{sec:bound}:
\begin{align}
  U M_V = \sum_\ell \op{\ell} \otimes M_{V\ell},
  &&
  M_{V\ell} = \sum_{m,\v v} D^\ell_{m,0} \op{m}{\v v}.
\end{align}
The block diagonal structure of this matrix implies that, rather than an integer index $k\in\set{1,2,\cdots,d^2}$, we can identify the singular values and right singular vectors of $UM_V$ (and equivalently $M_V$) by a block index $\ell\in\set{0,1,\cdots,d-1}$ and a sub-index $m=\set{-\ell,-\ell+1,\cdots,\ell}$.
Making the replacement $k\to\p{\ell,m}$, we can thus repeat the derivations through Eq.~\eqref{eq:bound} of Section \ref{sec:bound} without the assumption of Eq.~\eqref{eq:error_assumption}, finding that
\begin{align}
  \Bbk{\abs{\tilde\rho_{\ell m}^V-\rho_{\ell m}^V}^2}
  = \sum_{i,i'} q_{\ell m i}^V \p{q_{\ell m i'}^V}^*
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell}}
  = \sum_i \abs{q_{\ell m i}^V}^2
  \bbk{\epsilon_{i\ell}^2}
  \le \f{2\ell+1}{nd} \p{1-\delta_{\ell,0}}.
\end{align}
Altogether, we thus find that $n$ measurements of spin projection along every axis $\v v\in V$ allows us to construct an empirical estimate $\tilde\rho_V$ of an unknown state $\rho$ that is accurate to within a mean squared distance
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \sum_{\ell,m} \Bbk{\abs{\tilde\rho_{\ell m}^V-\rho_{\ell m}^V}^2}
  \le \f1{nd} \sum_{\ell>0} \p{2\ell+1} \S_{V\ell}^2,
\end{align}
where
\begin{align}
  \S_{V\ell} \equiv \norm{M_{V\ell}^{-1}}
  = \sqrt{\sum_m \p{M^V_{\ell m}}^{-2}}
\end{align}
is the (Hilbert-Schmidt, or Frobenius) norm of the right inverse of $M_{V\ell}$.

>>>>

Consider a $d$-state spin qudit with total spin $s\equiv\frac{d-1}{2}$.
The defining property of a spin qudit is the fact that it describes an angular momentum degree of freedom, which has specific implications for how a spin qudit should transform under the group SO(3) of rotations in 3D space.
Due to the central importance of these transformation rules for a spin qudit, we first identify a basis of operators that transforms nicely under 3D rotations.
One such basis is provided by the {\it transition operators} (also known as polarization operators \cite{kryszewski2006positivity, bertlmann2008bloch}), defined by
\begin{align}
  T_{\ell m} \equiv \sqrt{\f{2\ell+1}{2s+1}} \sum_{\mu,\nu=-s}^s
  \bk{s\mu;\ell m|s\nu} \op{\nu}{\mu},
  \label{eq:trans_op}
\end{align}
where $\bk{s\mu;\ell m|s\nu}$ is a Clebsh-Gordan coefficient that enforces $\ell\in\set{0,1,\cdots,d-1}$ and $m\in\set{-\ell,-\ell+1,\cdots,\ell}$, such that there are $d^2$ operators in total.
For brevity, we will generally treat the value of $d$ (or $s$) as constant but arbitrary throughout this work, and suppress any explicit dependence of quantities or operators such as $T_{\ell m}$ on $d$.
The transition operators are orthonormal with respect to the trace inner product, and transform nicely under conjugation:
\begin{align}
  \obk{T_{\ell m }|T_{\ell'm'}}
  = \delta_{\ell\ell'} \delta_{mm'},
  &&
  T_{\ell m}^\dag = \p{-1}^m T_{\ell,-m},
\end{align}
where for any $d\times d$ matrix $X=\sum_{\mu,\nu} X_{\mu\nu} \op{\mu}{\nu}$ we define the $d^2$-component vector $\oket{X} \equiv \sum_{\mu,\nu} X_{\mu\nu} \ket{\mu\nu}$; $\obra{X}$ is the conjugate transpose of $\oket{X}$; and $\delta_{kk'}\equiv 1$ if $k=k'$ and $0$ otherwise.
As a consequence, we can expand any density operator $\rho$ in the transition operator basis as
\begin{align}
  \rho = \sum_{\ell=0}^{2s} \sum_{m=-\ell}^\ell \rho_{\ell m} T_{\ell m},
  &&
  \rho_{\ell m} \equiv \obk{T_{\ell m}|\rho},
  \label{eq:trans_state}
\end{align}
where $\rho^\dag=\rho$ implies that $\rho_{\ell m}^*=\p{-1}^m\rho_{\ell,-m}$.
The transition operators can be physically interpreted as the transition $\ket\psi\to T_{\ell m}\ket\psi$ induced on the state $\ket\psi$ of a spin-$s$ system by the absorption of a spin-$\ell$ particle with spin projection $m$ onto a fixed quantization axis.
In analogy with the spherical harmonics $Y_{\ell m}$, we will refer to the index $\ell$ as the {\it degree} and $m$ as the {\it order} of $T_{\ell m}$.

@@

Rather than the statistical errors $\epsilon_{i\mu} \equiv \tilde\Pi_{i\mu} - \bk{\Pi_{i\mu}}_\rho$ in the estimates $\tilde\Pi_{i\mu}$ of the projectors $\Pi_{i\mu}$, we now consider the statistical errors $\epsilon_{i\ell} \equiv \tilde T_{i\ell} - \bk{T_{i\ell}}_\rho$ in the estimates $\tilde T_{i\ell}$ of the transition operators $T_{i\ell} \equiv T_{\v v_i\ell,0}$.
We can expand the transition operators $T_{i\ell}$ as a sum over projectors $\Pi_{i\mu}$ as
\begin{align}
  T_{i\ell} = \sum_\mu t_{\ell\mu} \Pi_{i\mu},
  &&
  t_{\ell\mu} \equiv \bk{\mu|T_{\ell,0}|\mu},
\end{align}
and likewise $\tilde T_{i\ell} = \sum_\mu t_{\ell\mu} \tilde\Pi_{i\mu}$.
The covariances between statistical errors in the transition operator basis are then
\begin{align}
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell'}}
  = \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \bbk{\epsilon_{i\mu} \epsilon_{i'\mu'}}
  = \delta_{ii'} \times \f1n \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell'\mu'}
  \sp{\delta_{\mu\mu'} p_\mu^{(i)} - p_\mu^{(i)} p_{\mu'}^{(i)}}.
\end{align}
In order to establish an upper bound on reconstruction error for an unknown quantum state, we need an upper bound on the variance $\bbk{\epsilon_{i\ell}^2}$ in particular.
To this end, for any multinomial probability distribution $p\equiv\p{p_1,p_2,\cdots,p_d}$ we define the function
\begin{align}
  g_\ell\p{p} \equiv \sum_{\mu,\mu'} t_{\ell\mu} t_{\ell\mu'}
  \p{\delta_{\mu\mu'} p_\mu - p_\mu p_{\mu'}}
  = \sum_\mu t_{\ell\mu}^2 p_\mu - \p{\sum_\mu t_{\ell\mu} p_\mu}^2
  = \sigma_p^2\p{t_\ell},
\end{align}
where $t_\ell \equiv \p{t_{\ell,-s},t_{\ell,-s+1},\cdots,t_{\ell s}}$ is a vector of all $t_{\ell\mu}$, and
\begin{align}
  \sigma_p^2\p{X}
  \equiv \sum_\mu p_\mu \p{X_\mu - \sum_\nu p_\nu X_\nu}^2
  = \sum_\mu p_\mu X_\mu^2 - \p{\sum_\mu p_\mu X_\mu}^2
\end{align}
is the variance of $X$ when sampled according to the probability distribution $p$.
By expanding
\begin{align}
  t_{\ell\mu} = \sqrt{\f{2\ell+1}{d}}\, \tau_{\ell\mu},
  &&
  \tau_{\ell\mu} \equiv \bk{s\mu;\ell,0|s\mu},
\end{align}
we can simplify
\begin{align}
  g_\ell\p{p} = \f{2\ell+1}{d} \times \sigma_p^2\p{\tau_\ell}.
\end{align}
We now seek to find an upper bound for $g_\ell\p{p}$, which will provide an upper bound for $\bbk{\epsilon_{i\ell}^2}$.
To bound $g_\ell\p{p}$, we note that variance $\sigma_p^2\p{X}$ is maximal when $p$ has equal weight on the largest and smallest values of $X$, which implies that
\begin{align}
  \sigma_p^2\p{\tau_\ell} \le \gamma_\ell^2,
  &&
  \gamma_\ell \equiv \f{\max\tau_\ell - \min\tau_\ell}{2}.
\end{align}
Altogether, using additionally the fact that $g_0\p{p}=0$, we can bound
\begin{align}
  \bbk{\epsilon_{i\ell}^2}
  = \f1n \times g_\ell\p{p^{(i)}}
  \le \f{2\ell+1}{nd} \times \gamma_\ell^2
  \times \p{1 - \delta_{\ell,0}}.
  \label{eq:variance_bound}
\end{align}
Note that the bound in Eq.~\eqref{eq:variance_bound} is tight, as equality is achieved by the mixed state
\begin{align}
  \rho^\star
  \equiv \f12 \p{\Pi_{\v v_i,\mu_{\t{max}}} + \Pi_{\v v_i,\mu_{\t{min}}}},
\end{align}
where $\mu_{\t{max}}$ ($\mu_{\t{min}}$) is the index that maximizes (minimizes) $\tau_{\ell\mu}$.
In addition, the  Clebsch-Gordan coefficients $\tau_{\ell\mu}$ are at most $1$ in magnitude, which implies that $\gamma_\ell^2\le 1$.
However, this bound on $\gamma_\ell^2$ is not tight in general, as we could have alternately bounded
\begin{align}
  g_\ell\p{p} \le \sum_\mu t_{\ell\mu}^2 p_\mu
  \le \sum_\mu t_{\ell\mu}^2 = 1,
\end{align}
which must also hold for the probability distribution $p^\star$ that maximizes $g_\ell$, i.e.
\begin{align}
  g_\ell\p{p^\star} = \f{2\ell+1}{d} \times \gamma_\ell^2 \le 1,
  &&
  \t{so}
  &&
  \gamma_\ell^2 \le \min\Set{1, \f{d}{2\ell+1}}.
\end{align}

@@

Equipped with the bound on variances in Eq.~\eqref{eq:variance_bound}, we again consider the block-diagonalized form of the measurement matrix derived in Section \ref{sec:bound}:
\begin{align}
  U M_V = \sum_\ell \op{\ell} \otimes M_{V\ell},
  &&
  M_{V\ell} = \sum_{m,\v v} D^\ell_{m,0} \op{m}{\v v}.
\end{align}
The block diagonal structure of this matrix implies that, rather than an integer index $k\in\set{1,2,\cdots,d^2}$, we can identify the singular values and right singular vectors of $UM_V$ (and equivalently $M_V$) by a block index $\ell\in\set{0,1,\cdots,d-1}$ and a sub-index $m=\set{-\ell,-\ell+1,\cdots,\ell}$.
Making the replacement $k\to\p{\ell,m}$, we can thus repeat the derivations through Eq.~\eqref{eq:bound} of Section \ref{sec:bound} without the assumption of Eq.~\eqref{eq:error_assumption}, finding that
\begin{align}
  \Bbk{\abs{\tilde\rho_{\ell m}^V-\rho_{\ell m}^V}^2}
  = \sum_{i,i'} q_{\ell m i}^V \p{q_{\ell m i'}^V}^*
  \bbk{\epsilon_{i\ell} \epsilon_{i'\ell}}
  = \sum_i \abs{q_{\ell m i}^V}^2
  \bbk{\epsilon_{i\ell}^2}
  \le \f{2\ell+1}{nd} \times \gamma_\ell^2 \times \p{1-\delta_{\ell,0}}.
\end{align}
Altogether, we thus find that $n$ measurements of spin projection along every axis $\v v\in V$ allows us to construct an empirical estimate $\tilde\rho_V$ of an unknown state $\rho$ that is accurate to within a mean squared distance
\begin{align}
  \Bbk{\norm{\tilde\rho_V-\rho}^2}
  = \sum_{\ell,m} \Bbk{\abs{\tilde\rho_{\ell m}^V-\rho_{\ell m}^V}^2}
  \le \f1{nd} \sum_{\ell>0} \p{2\ell+1} \gamma_\ell^2 \S_{V\ell}^2,
\end{align}
where
\begin{align}
  \S_{V\ell} \equiv \norm{M_{V\ell}^{-1}}
  = \sqrt{\sum_m \p{M^V_{\ell m}}^{-2}}
\end{align}
is the (Hilbert-Schmidt, or Frobenius) norm of the right inverse of $M_{V\ell}$.
